An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-05 main.tex
Work on introduction

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  A common problem in string constraint solvers is computing the Parikh image, a
   set of linear equations that describe all possible combinations of character
   counts in strings of a given language. Automata-based string solvers
   frequently need to compute the Parikh image of large products (intersections)
   of automata, which in many operations is both prohibitively slow and
   memory-intensive. We contribute a novel understanding of how Parikh maps can
   be combined with arbitrary commutative monoid morphisms to represent
   real-world constraints on regular languages, most notably the length
   constraint. Furthermore, we show how this formulation can be efficiently
   implemented as a calculus, \Calculus{}, in an automated theorem prover
   supporting Presburger logic. We also provide an efficient method for deriving
   the Presburger formula of a given Parikh image, \Fudge{which is good for
   something} based on techniques for quantifier elimination. We implement
   \Calculus{} in a tool called \Catra{}, and evaluate it on constraints
   produced by the \Ostrich{} string constraint solver when solving
   \Fudge{standard string constraint benchmarks involving string lengths}. We
   show that our solution strictly outperforms the standard eager approach
   described in~\citeauthor{generate-parikh-image} as well as the
   over-approximating method recently described
   by~\citeauthor{approximate-parikh}, and in many cases also the~\Nuxmv{} model
   checker.
\end{abstract}

@@

The Parikh map is a characterisation of a context-free language as a vector of
its character counts, famously a semilinear set~\cite{parikh-theorem}. It
appears naturally as part of \Fudge{many operations} in model checking and
solving string constraints in automata-based solvers such as \Ostrich{}
\cite{ostrich}, notably in representing constraints on string lengths.
Specialised string solvers like e.g. Trau \cite{trau} also perform extensive
reasoning using Parikh images, but do so on restricted automata that do not
suffer from the blow-up we seek to address, though it might be possible to
adapts aspects of our approach to theirs.

@@

In addition to the elimination of existentially quantified variables, another
problem arises from length constraints in string solvers. Conjunctions of string
constraints lead to the computation of length images of products (intersections)
of regular languages represented as finite automata. Applying the approach
in~\cite{generate-parikh-image} would require the eager computation of the
product before its length image. In several instances we have observed while
solving real-world string constraints, the computation of the product of
automata exhausts the memory of any machine available due to the exponential
blow-up in size of the product, quickly becoming intractable as the number of
automata in the product increases. The current best published mitigation for
this problem is an over-approximation that works by approximating the Parikh
image of a product of automata to be the conjunction of the image of the
individual automata of the product \cite{approximate-parikh}. This approach only
works for unsatisfiable instances, and comes with a harsh penalty on satisfiable
instances.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary Parikh-like
homomorphism over automata labels, including string lengths. This enables us to
let both calculations inform each other, eliminating unnecessary work, and
pruning the size of the partial products considered in the computation for a
smaller memory footprint. Moreover, the method can be used iteratively to tackle
smaller chunks of the product incrementally, thereby avoiding running out of
memory at all.

@@

We implement \Calculus{} as a theory for the \Princess{} automated theorem
prover in a tool that we call \Catra{}. \Catra also supports using the
approximate method of \cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra{}, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} Parikh automata satisfaction problems generated
\Fudge{by Ostrich+}.

@@

@@

To illustrate the key features of \Calculus, we start with two regular
languages, a commutative homomorphism $f$ on them that we care about, and a side
constraint for which we want to find values in $f$'s image over the intersection
of the two languages using a heavily elided version of \Calculus. We will use
the same two regular languages, $\mathtt{ac^*a|b(bb)^*c}$ and $\mathtt{.^*c.^*}$
for both examples but change the constraint $\varphi$ and homomorphism $f$. In
essence, we solve the following set of constraints for all possible values of
$s$ and $\MonoidElement$:

\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{ac^*a|b(bb)^*c} \text{ accepts } s\\
  & \mathtt{.^*c.^*} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

We will represent the regular languages as the automata $\SomethingCSomething{}$
 (\cref{fig:something-c-something}) and $\AcaOrBc{}$ (\cref{fig:aca-or-bc})
 respectively.

@@

To keep things simple, in this first example we
let $f$ be the function that gives the length of a given string, e.g.
$f\left(\text{curious and curiouser still}\right) = 27$, and $f\left(\text{homo}\right) +
f\left(\text{morphism}\right) = f\left(\text{homomorphism}\right) = 12$.

\begin{figure}[ht]
  \begin{minipage}[b]{0.75\linewidth}
  \centering 
    \includegraphics[width=\textwidth]{aca_or_bc}
    \caption{An automaton recognising the regular expression
    $\mathtt{ac^*a|b(bb)^*c}$.}\label{fig:aca-or-bc}

    \includegraphics[width=\textwidth]{something_c_something}
    \caption{An automaton recognising the regular expression
    $\mathtt{.^*c.^*}$.}\label{fig:something-c-something}
  \end{minipage}
  \end{figure}

@@

To give an intuition for the core idea of \Calculus{}, we will annotate the
transitions of $\SomethingCSomething$ and $\AcaOrBc$ with variables illustrating
how many times they are taken, and perform informal reasoning on them
corresponding to the reasoning performed by our calculus.

\begin{figure}[ht]
  \centering 
    \includegraphics[width=0.75\linewidth]{aca_or_bc_simplified}
    \caption{$\AcaOrBc{}$ with its symbolic transition counters after
    simplification of their equations.}\label{fig:example-simplify-1}
  \end{figure}

Note that we have already performed some reasoning in the starting figures; we
use the same symbol for the number of times each of the upper transitions of
$\AcaOrBc$ are taken, since we know that they must be taken the same number of
times, and we introduced the constant $1$ for the middle transition in
$\SomethingCSomething$, since it must always be taken precisely once to reach
the accepting state. We use standard graph-theoretical flow reasoning to set the
initial flow into the starting state to 1 and use linear equations to enforce
balanced flow throughout the automaton to determine the relations between the
counting variables on each transition (e.g. $1 + {l_b}' = l_a + l_b$, or $l_a +
l_c' = 1$). Strictly speaking, this is where the constant number $1$ in
\cref{fig:something-c-something} comes: from $1 + r_\Sigma = x + r_\Sigma$
solved for $x$. Using analogous linear equation reasoning loses additional
variables and gives us the annotations in \cref{fig:example-simplify-1}.

We sum the individual transition variables up to get the length we are actually
interested in, e.g. the image of $f$ on $\SomethingCSomething =
\exists_{r_\Sigma, r_\Sigma'} \: r_\Sigma + 1 + r_\Sigma'$. For
$\SomethingCSomething{}$, this is all we need to produce the image since all
transitions are always accessible from any path from the initial state to the
accepting state.

The story gets more complicated with $\AcaOrBc{}$, where we have a choice of two
paths, and where that choice affects the possibility of taking either of the two
loops. In that case (letting the existential quantifiers from now on be
implicit), we would have the image $2l_a + l_c + l_c' + l_b + l_b'$. If we apply
standard linear reasoning on the equations (e.g. $1 + {l_b}' = l_a + l_b$, or
$l_a + l_c' = 1$), we can reduce the number of bound ($\exists$-quantified)
variables somewhat, obtaining the relations in
Figure~\ref{fig:example-simplify-1}, and the corresponding over-approximation on
the $f$-image $2 + l_c + 2l_b'$. It is an over-approximation because it only
considers the number of times each transition is taken and the existence of a
path from an initial to an accepting state, but omits the connectivity of that
path in the presence of loops. For example, no valid path would have both $l_c >
0$ and $l_b' > 0$, but no such restriction exists in the formula above.

\begin{figure}[ht]
  \centering 
    \includegraphics[width=0.75\linewidth]{something_c_something_length_simplified}
    \caption{$\SomethingCSomething{}$ after using linear reasoning over lengths to replace a variable.}\label{fig:example-length-reduced}
  \end{figure}

Still, this approximate solution allows us to perform further reasoning. Since
we know that the length of a string accepted by the intersection of the two
automata's languages must be the same, we also know that their images (string
lengths) must be the same under $f$, e.g. that $r_\Sigma + 1 + r_\Sigma' = 2 +
l_c + 2l_b' \iff r_\Sigma + r_\Sigma' = 1 + l_c + 2l_b'$. This in turn allows us
to replace one additional variable, e.g. $r_\Sigma = 1 + l_c + 2l_b' -
r_\Sigma'$ (see Figure~\ref{fig:example-length-reduced}). Note that the two
automata are now in direct contact. Any reasoning performed on one of them could
be propagated to the other.

@@

We talked initially about having a constraint on the image (or conversely, on
the strings of the language), so let us add one. We are only interested in
strings of odd length, e.g. $2 \nmid  f(\AcaOrBc \times \SomethingCSomething)$.
This means that: $2 \nmid 2 + l_c + 2l_b' \iff 2 \nmid l_c$. To encode this condition, we simply replace $l_c$ with $2k + 1$ with an existentially quantified integer $k \geq 0$.

This is as far as we can get with this reasoning. We must now choose: either we
compute the product and continue our reasoning, or we perform a case split. Let
us try a case split in order to put off computing a potentially large product.
We pick an early transition in the largest automaton with loops, in our case
$\FromLabelTo{start}{a}{sawA}$ of $\AcaOrBc$. This creates two cases: $1-l_c' =
0$ and $1-l_c' > 0$. We start with the first case, and after propagating the now known values and removing any transition that becomes zero, we get the automaton of Figure~\ref{fig:aca-or-bc-length-split-1}.

\begin{figure}[ht]
  \centering 
    \includegraphics[width=0.75\linewidth]{aca_or_bc_length_split_1}
    \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{start}{a}{sawA}$ is not used.}\label{fig:aca-or-bc-length-split-1}
  \end{figure}

  We immediately notice that the self-loop of state $sawA$ is now disconnected
  and must be removed. This condition can be detected efficiently using standard
  forwards/backwards reachability from the initial and accepting states
  respectively. In other words, $2k+1 = 0$ for this case and is implied by the splitting constraint. However, no choice of integer $k$  will satisfy this equation. Therefore, we close this branch and proceed with the other one -- where $1-l_c' > 0$. Note that we do so without ever computing a potentially expensive product of automata.

  \begin{figure}[ht]
    \centering 
      \includegraphics[width=0.75\linewidth]{aca_or_bc_length_split_2}
      \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{start}{a}{sawA}$ is used.}\label{fig:aca-or-bc-length-split-2}
    \end{figure}

Plugging in the assumption $1-l_c' > 0$ (and its consequence that $l_c' = 0$),
we also get a reduced variant of $\AcaOrBc$, as seen in
Figure~\ref{fig:aca-or-bc-length-split-1}. Note that $sawB$ is no longer
backwards-reachable from $final$ and therefore dead, along with both transitions
into it. We add the conclusion $l_b' = 0$, which severs both transitions into
the state and removes it from our automaton. Following this we are ready to
compute the product, which can be seen in Figure~\ref{fig:length-split-product}.

\begin{figure}[ht]
  \centering 
    \includegraphics[width=0.75\linewidth]{length_split_product}
    \caption{The product of $\AcaOrBc$ and $\SomethingCSomething$ after
    case splitting and removal of dead states.}\label{fig:length-split-product}
  \end{figure}

@@

  We now need to relate the number of times a transition was taken to the
  character counts we are really interested in, as opposed to just the length in
  the previous example. Note that this, like the length calculus before, puts
  the two automata into contact with each other, before any product has been
  computed. Some consideration leads us to the definition that the number of
  occurrence of each character corresponds to the sum of usages of transitions
  containing it, e.g. relations of the kind $\TransitionCount{a} = 2 - 2l_c$,
  and $\TransitionCount{b} = 2 l_b' + l_c'$.\footnote{For the attentive reader:
  in this example, we leave the catch-all transitions of $\SomethingCSomething$
  unconstrained since they will be taken care of by the product and would
  complicate the reasoning since we cannot yet know which character they will
  correspond to in the final product. In other words, this is a safe
  under-approximation that only limits our ability to reason before computing
  the product.}

This relation can be combined with the constraint that $\TransitionCount{a} >
\TransitionCount{b}$, that is that there are more instances in the accepted
string of letters a than b, to obtain $2 - 2l_c > 2 l_b' + l_c'$. Combining this
equation with the requirement from $\SomethingCSomething$ that
$\TransitionCount{c} \geq 1$ and $\TransitionCount{c} = l_c + l_c'$ from
$\AcaOrBc$ gives $0 \geq l_b'$, and since we cannot use a transition a negative
number of times $l_b' = 0$, which gives the automaton in
Figure~\ref{fig:example-simplify-2}.

@@

To obtain the final Parikh image, we need to tie the inequalities for the
transitions of the automata of the product to the new transition counters we
introduced in the product automaton. The only interesting transition is the one
carrying the c, so we will choose that one for illustration. In that case, the
bridging equation is $r_2 = l_c \land r_2 = r_{\Sigma}$ (and for all others
$=1$), which gives the Parikh image $\exists_{r_2} \TransitionCount{a} = 1 \land
\TransitionCount{b} = 0 \land \TransitionCount{c} = r_2 + 1$, and after trivial
existence-elimination using reasoning on lower bounds (e.g. $r_2 \geq 0$) we
arrive at $\TransitionCount{c} \geq 1$. For comparison, the $\AcaOrBc{} \times
\SomethingCSomething{}$ we would have obtained if we had eagerly computed the
product can be seen in Figure~\ref{fig:original-product}. This would have
introduced a significantly more complicated existentially quantified formula.

@@

\subsection{Finite-state Automata and their Products}
We define a non-deterministic automaton~$\Automaton$ with alphabet~$\Alphabet$
as $\Automaton = \AutomatonTuple$ where $\Transitions = \States \times \Alphabet
\times \States$, $\States$ is its states, $\InitialState$ is w.l.o.g. assumed to
be the single initial state, and $\AcceptingStates$ is its set of accepting
states.  We describe a transition $\Transition$ from state $\State$ to state
$\State'$ with label $\Label \in \Alphabet$ as $\Transition =
\FromLabelTo{\State}{\Label}{\State'}$. When one or more of the states and
labels of a transition are uninteresting we will omit it.

@@

\subsection{The Parikh Map and its Image}
Formally, the \textit{Parikh map} over a context-free language $\Strings=
\left\{a_1, \ldots, a_k \right\}$ is defined as in \cite{kozen}:

$$
\begin{aligned}
& \ParikhMap: \MapFromTo{\Strings}{\natural^k} \\
& \ParikhMap(s) = \VectorLiteral{\#a_1(s), \#a_2(s), \ldots, \#a_k(s)}
\end{aligned}
$$

That is, $\ParikhMap(s)$ is a vector of the number of occurrences of each
character in the language for a given string $s$. For example, for  $\Strings =
\Set{a, b}$, we would have $\ParikhMap(abb) = \VectorLiteral{1, 2}$.

We define the image of this map, the \textit{Parikh image}, of some subset of
the language $\Language \subseteq \Strings$ as:

\[
\ParikhMap(\Language) = \Set{\ParikhMap(x) \SuchThat x \in \Language}
\]

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$. We also sometimes use the standard
notation $\#l$ to talk about an individual letter $l$. For example, for the Parikh vector above, we would have $\CountOf{a} = 1$. You have already seen 
this in Section~\ref{sec:introduction:parikh}.

Parikh's theorem states that ignoring order, any context-free grammar has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars and~\cite{bounds} for bounds on its size). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \CountOf{a} =
\CountOf{b} = \CountOf{c} \land \CountOf{a} \geq 0$).

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  A common problem in string constraint solvers is computing the Parikh image, a
   set of linear equations that describe all possible combinations of character
   counts in strings of a given language. Automata-based string solvers
   frequently need to compute the Parikh image of large products (intersections)
   of automata, which in many operations is both prohibitively slow and
   memory-intensive. We contribute a novel understanding of how Parikh maps can
   be tackled as a constraint-solving problem to solve real-world constraints
   stemming from functions on regular languages, most notably the length
   constraint. Furthermore, we show how this formulation can be efficiently
   implemented as a calculus, \Calculus{}, in an automated theorem prover
   supporting Presburger logic. We also provide an efficient method for deriving
   the Presburger formula of a given Parikh image, \Fudge{which is good for
   something} based on techniques for quantifier elimination. We implement
   \Calculus{} in a tool called \Catra{}, and evaluate it on constraints
   produced by the \OstrichPlus{} string constraint solver when solving
   \Fudge{standard string constraint benchmarks involving string lengths}. We
   show that our solution strictly outperforms the standard eager approach
   described in~\citeauthor{generate-parikh-image} as well as the
   over-approximating method recently described
   by~\citeauthor{approximate-parikh}, and for realistic timeouts for constraint
   solving also the~\Nuxmv{} model checker.
\end{abstract}

@@

The Parikh map is a characterisation of a context-free language as a vector of
its character counts, famously a semilinear set~\cite{parikh-theorem}. It
appears naturally as part of \Fudge{many operations} in model checking and
solving string constraints in automata-based solvers such as \Ostrich{}
\cite{ostrich}, notably in representing constraints on string lengths. The
enhanced \OstrichPlus{} solver~\cite{ostrich-plus} makes even more extensive use
of Parikh images.

@@

In addition to the elimination of existentially quantified variables, another
problem arises in string solvers in particular. Conjunctions of string
constraints lead to the computation of length images of products (intersections)
of regular languages represented as finite automata. Applying the approach
in~\cite{generate-parikh-image} would in this case require the eager computation
of the product before its length image. In several instances we have observed
while solving real-world string constraints, the computation of the product of
automata exhausts the memory of any machine available due to the exponential
blow-up in size of the product, quickly becoming intractable as the number of
automata in the product increases. The current best published mitigation for
this problem is an over-approximation that works by approximating the Parikh
image of a product of automata to be the conjunction of the image of the
individual automata of the product \cite{approximate-parikh}. This approach only
works for unsatisfiable instances, and comes with a harsh penalty for
satisfiable instances.

Addressing these concerns, we have developed a calculus for Parikh images of
products of regular languages that we call \Calculus{}. It allows us to
interleave the computation of arbitrarily deep products of automata with the
product's Parikh image, and is generalised to an arbitrary homomorphism over
automata labels, including string lengths. This enables us to let both
calculations inform each other, eliminating unnecessary work, and pruning the
size of the partial products considered in the computation for a smaller memory
footprint. Moreover, the method can be used iteratively to tackle smaller chunks
of the product incrementally, thereby avoiding running out of memory at all.

@@

We implement \Calculus{} as a theory for the \Princess{} automated theorem
prover in a tool that we call \Catra{}. \Catra also supports using the
approximate method of \cite{approximate-parikh}, its fall-back variant adapted
from~\cite{generate-parikh-image}, and an adapter for the \Nuxmv{} model
checker~\cite{nuxmv}. Using \Catra{}, we compare \Calculus{} to the other two
back-ends on \NrBenchmarks{} Parikh automata satisfaction problems generated
\Fudge{by \OstrichPlus{}}.

@@

\subsection{Related Work}

Parikh image computations as well as Parikh automata \cite{parikh-automata}
feature extensively in string solvers, including as mentioned above \Ostrich{}
and \OstrichPlus{} \cite{ostrich,ostrich-plus}, but also formes the basis of
Trau~\cite{trau-pldi}, and occurs in \textsc{Sloth}~\cite{sloth}. Parikh images
frequently appear when introducing cardinality constraints like length or string
indexing. The state-of-the-art approach to handling Parikh image computation is
to over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Automaton_1 \times \ldots \times \Automaton_k)$ with the
conjunction of the automata's parikh maps, $\AndComp{i=1}^{k}
\ParikhMap(\Automaton_i)$. This approach works only for unsatisfiable instances,
and will require falling back to computing the product of the automata before
using the standard approach for finding its image originally presented in~\cite{generate-parikh-image}.

Outside of string solvers, Parikh automata have been proposed as the basis of
queries~\cite{graph-queries}, and for solving cardinalities in model checking
problems involving epistemic logic~\cite{epistemic-logic}.

@@

To illustrate the key features of \Calculus, we will informally present our
reasoning. We begin by considering the following set of string and other
constraints, which we want to solve for all possible values of $\MonoidElement$,
a value in some commutative monoid \Fudge{MONOID} and the string $s$, where
$\varphi(\MonoidElement)$ is some set of constraints on $\MonoidElement$:
\begin{equation*}
  \AndComp{}{\begin{aligned}
  & \mathtt{ac^*a|b(bb)^*c} \text{ accepts } s\\
  & \mathtt{.^*c.^*} \text{ accepts } s \\
  & \MonoidElement = f(s) \\
  & \varphi(\MonoidElement)
  \end{aligned}}
 \end{equation*}

We will represent the strings using the nondeterministic finite automata
 $\SomethingCSomething{}$ (\cref{fig:something-c-something}) and $\AcaOrBc{}$
 (\cref{fig:aca-or-bc}) respectively.

 Throughout these examples we will present our three key ideas:
 \emph{automata-aware splitting}, \emph{lazy enforcement of automata
 connectivity}, and \emph{incremental materialisation of products}.

@@

To keep things simple, in this first example we let $f$ be the function that
gives the length of a given string, e.g. $f\left(\text{curious and curiouser
still}\right) = 27$, and $f\left(\text{homo}\right) +
f\left(\text{morphism}\right) = f\left(\text{homomorphism}\right) = 12$, and we
let $\varphi(\MonoidElement) = \exists k \HoldsThat m = 2k + 1 \land k \geq 0$,
i.e. require that the length of the string is odd.

\begin{figure}[t]
  \begin{minipage}[b]{0.50\linewidth}
  \centering 
    \includegraphics[width=\textwidth]{aca_or_bc}
    \caption{An automaton recognising the regular expression
    $\mathtt{ac^*a|b(bb)^*c}$.}\label{fig:aca-or-bc}

    \includegraphics[width=\textwidth]{something_c_something}
    \caption{An automaton recognising the regular expression
    $\mathtt{.^*c.^*}$.}\label{fig:something-c-something}
  \end{minipage}
  \Description[Two deterministic finite automata, one above the other]{NFAs 
    recognising the languages ac^*a|b(bb)^*c and ac^*a|b(bb)^*c respectively.
    The first one has 4 states and the second one 2.}
  \end{figure}

@@

Similar to \cite{generate-parikh-image} we associate each transition
$\Transition$ with an integer variable $\TransitionVar_\Transition \geq 0$
representing how many times $\Transition$ is used that we leave implicitly
existentially quantified. These variables are shown as annotations under the
transition labels in \cref{fig:aca-or-bc,fig:something-c-something}. We then
apply graph-like flow reasoning by adding linear equations (e.g. $1 + {l_b}' =
l_a + l_b$, or $l_a + l_c' = 1$) that equate the incoming transitions into a
state to the ones going out. For the initial and accepting states we add an
incoming and outgoing flow of 1, respectively. We will use these linear
equations to perform reasoning to reduce the size of the product before we
compute it. Note that we have already performed some linear reasoning to
eliminate some of the variables in the graphs above, for example to introduce
the constant number $1$ in \cref{fig:something-c-something}, which comes from $1
+ r_\Sigma = x + r_\Sigma$ solved for $x$.

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.50\linewidth]{aca_or_bc_simplified}
    \caption{$\AcaOrBc{}$ with its symbolic transition counters after
    simplification of their equations.}\label{fig:example-simplify-1}
    \Description[The automaton $\AcaOrBc{}$ labelled with variables counting the number of times each transition is taken, after having simplified the equations.]{The starting state shows an initial flow of 1, with two outgoing transitions to $A$ ($1-l_c'$ on letter a, and $l_b' + l_c'$ on b respectively). $A$ has a flow of $l_c$, and its outgoing transition into the final state, $F$ is the same as the incoming one from $S$, the initial state. The $B$ transition has an outgoing transition labeled $c/ l_c'$ to the final state $F$, and one backwards transition $b / l_b'$ to the starting state.}
  \end{figure}

We sum the individual transition variables to get the length $\MonoidElement$ we
are actually interested in. For example the image of $f$ on
$\SomethingCSomething$ is given by $\exists_{r_\Sigma, r_\Sigma'} \: 2k + 1 =
r_\Sigma + 1 + r_\Sigma' \land k \geq 0$. For $\SomethingCSomething{}$, this is
all we need to produce the image since all transitions are always accessible
from any path from the initial state to the accepting state. The difficult
connectivity constraint is not necessary to enforce here, since any path through
$\SomethingCSomething{}$ is connected, and therefore the flow equations
described above are sufficient to capture its image.

The story gets more complicated with $\AcaOrBc{}$, where we have a choice of two
paths, up or down, and where that choice affects the possibility of taking
either of the two loops. In that case, we would have the image $2l_a + l_c + l_c' + l_b + l_b'$.
If we apply standard linear reasoning on the equations (e.g. $1 + {l_b}' = l_a +
l_b$, or $l_a + l_c' = 1$), we can reduce the number of bound
($\exists$-quantified) variables somewhat, obtaining the relations in
\cref{fig:example-simplify-1}, and the corresponding over-approximation on
the $f$-image $2 + l_c + 2l_b'$. It is an over-approximation because it only
considers the number of times each transition is taken and the existence of a
path from an initial to an accepting state, but omits the connectivity of that
path in the presence of loops. For example, no valid path would have both $l_c >
0$ and $l_b' > 0$, but no such restriction exists in the formula above.

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.4\linewidth]{something_c_something_length_simplified}
    \caption{$\SomethingCSomething{}$ after using linear reasoning over lengths to replace a variable.}\label{fig:example-length-reduced}
    \Description[The automaton $\SomethingCSomething{}$ after reasoning over lengths.]{The same automaton as before for $\SomethingCSomething{}$, but with the transition from start to final annotated with a constant number 1, and the leftmost self-loop transition replaced with the equation $1 + l_c + 2l_b' -
    r_\Sigma'$.}
  \end{figure}

Still, this approximate solution allows us to perform further reasoning. Since
we know that the length of a string accepted by the intersection of the two
automata's languages must be the same, we also know that their images (string
lengths) must be the same under $f$, e.g. that $r_\Sigma + 1 + r_\Sigma' = 2 +
l_c + 2l_b' \iff r_\Sigma + r_\Sigma' = 1 + l_c + 2l_b'$. This in turn allows us
to replace one additional variable, e.g. $r_\Sigma = 1 + l_c + 2l_b' -
r_\Sigma'$ (see \cref{fig:example-length-reduced}). Note that the two automata
are now in contact, allowing us to propagate reasoning on one to the other.

@@

This is as far as we can get with this reasoning. We must now choose: either we
compute (materialise) the product and continue our reasoning, or we perform a
case split. Let us try a case split in order to put off computing a potentially
large product. 1 an early transition in the largest automaton with loops,
in our case $\FromLabelTo{S}{a}{A}$ of $\AcaOrBc$. This creates two
cases: $1-l_c' = 0$ and $1-l_c' > 0$. We start with the first case, and after
propagating the now known values and removing any transition that becomes zero,
we get the automaton of \cref{fig:aca-or-bc-length-split-1}.

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.75\linewidth]{aca_or_bc_length_split_1}
    \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{S}{a}{A}$ is not
    used.}\label{fig:aca-or-bc-length-split-1}
    \Description[The automaton \AcaOrBc{} where we have removed the transition
    from $S$ to $A$.]{ The state $A$ hangs free without incoming transitions,
    and only the lower part of the automaton remains.}
  \end{figure}

  We immediately notice that the self-loop of state $A$ is now disconnected and
  must be removed. This condition can be detected efficiently using standard
  forwards/backwards reachability from the initial and accepting states
  respectively. In other words, $2k+1 = 0$ for this case and is implied by the
  splitting constraint. However, no choice of integer $k$  will satisfy this
  equation. Therefore, we close this branch and proceed with the other one --
  where $1-l_c' > 0$. Note that we do so without ever computing a potentially
  expensive product of automata.

  \begin{figure}[t]
    \centering 
      \includegraphics[width=0.5\linewidth]{aca_or_bc_length_split_2}
      \caption{$\AcaOrBc{}$ at the split where $\FromLabelTo{S}{a}{A}$ \emph{is} used.}\label{fig:aca-or-bc-length-split-2}
      \Description[$\AcaOrBc{}$ with the transition from $B$ to $F$ removed.]{$\AcaOrBc{}$ when we keep the transition from $S$ to $A$. This removes the transition between $B$ and $F$, but otherwise maintains the automaton.}
    \end{figure}

Plugging in the assumption $1-l_c' > 0$ (and its consequence that $l_c' = 0$),
we also get a reduced variant of $\AcaOrBc$, as seen in
\cref{fig:aca-or-bc-length-split-2}. Note that there is now no way to select any
transition variable so as to disconnect a loop from a path between $S$ and $F$,
and we can therefore stop worrying about enforcing connectivity.

\begin{figure}[t]
  \centering 
    \includegraphics[width=0.5\linewidth]{length_split_product}
    \caption{The product of $\AcaOrBc$ and $\SomethingCSomething$ after
    case splitting and removal of dead states.}\label{fig:length-split-product}
  \end{figure}

@@

  We now need to relate the number of times a transition was taken to the
  character counts we are really interested in, as opposed to just the length in
  the previous example. Note that this, like the length calculus before, puts
  the two automata into contact with each other, before any product has been
  computed. Some consideration leads us to the definition that the number of
  occurrence of each character corresponds to the sum of usages of transitions
  containing it, e.g. relations of the kind $\TransitionCount{a} = 2 - 2l_c'$,
  and $\TransitionCount{b} = 2 l_b' + l_c'$.\footnote{For the attentive reader:
  in this example, we leave the catch-all transitions of $\SomethingCSomething$
  unconstrained since they will be taken care of by the product and would
  complicate the reasoning since we cannot yet know which character they will
  correspond to in the final product. In other words, this is a safe
  under-approximation that only limits our ability to reason before computing
  the product.}

This relation can be combined with the constraint that $\TransitionCount{a} >
\TransitionCount{b}$, that is that there are more instances in the accepted
string of letters a than b, to obtain $2 - 2l_c' > 2 l_b' + l_c'$. Combining this
equation with the requirement from $\SomethingCSomething$ that
$\TransitionCount{c} \geq 1$ and $\TransitionCount{c} = l_c + l_c'$ from
$\AcaOrBc$ gives $0 \geq l_b'$, and since we cannot use a transition a negative
number of times $l_b' = 0$, which gives the automaton in
Figure~\ref{fig:example-simplify-2}.

@@

To obtain the final Parikh image, we need to tie the inequalities for the
transitions of the automata of the product to the new transition counters we
introduced in the product automaton. The only interesting transition is the one
carrying the c, so we will choose that one for illustration. In that case, the
bridging equation is $r_2 = l_c \land r_2 = r_{\Sigma}$ (and for all others
$=1$), which gives the Parikh image $\exists_{r_2} \TransitionCount{a} = 1 \land
\TransitionCount{b} = 0 \land \TransitionCount{c} = r_2 + 1$, and after
existence-elimination using reasoning on lower bounds (e.g. $r_2 \geq 0$) we
arrive at $\TransitionCount{c} \geq 1$. For comparison, the $\AcaOrBc{} \times
\SomethingCSomething{}$ we would have obtained if we had eagerly computed the
product can be seen in Figure~\ref{fig:original-product}. This would have
introduced a significantly more complicated existentially quantified formula.

@@

\contents{\item Monoid
\item Alfabet
\item ord oever alfabet
\item konkatenering
}

\subsection{Finite-state Automata and their Products}
We define a non-deterministic automaton~$\Automaton$ with alphabet~$\Alphabet$
as $\Automaton = \AutomatonTuple$ where $\Transitions = \States \times \Alphabet
\times \States$, $\States$ is its states, $\InitialState$ is w.l.o.g. assumed to
be the single initial state, and $\AcceptingStates$ is its set of accepting
states.  We describe a transition $\Transition$ from state $\State$ to state
$\State'$ with label $\Label \in \Alphabet$ as $\Transition =
\FromLabelTo{\State}{\Label}{\State'}$. \Fudge{When one or more of the states and
labels of a transition are uninteresting we will omit them}.

@@

\subsection{The Parikh Map and its Image}
Formally, the \textit{Parikh map} over an alphabet $\Alphabet=
\left\{a_1, \ldots, a_k \right\}$ is defined as in \cite{kozen}:
$$
\begin{aligned}
& \ParikhMap: \MapFromTo{\Strings}{\natural^k} \\
& \ParikhMap(s) = \VectorLiteral{\#a_1(s), \#a_2(s), \ldots, \#a_k(s)}
\end{aligned}
$$

That is, $\ParikhMap(s)$ is a vector of the number of occurrences of each
character in the language for a given string $s$. For example, for  $\Alphabet =
\Set{a, b}$, we would have $\ParikhMap(abb) = \VectorLiteral{1, 2}$.

We define the image of this map, the \textit{Parikh image}, of some subset of
the language $\Language \subseteq \Strings$ as:
\[
\ParikhMap(\Language) = \Set{\ParikhMap(x) \SuchThat x \in \Language}
\]

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$. We also sometimes use the standard
notation $\#l(w)$ to talk about an individual letter $l$ in a word $w$. For
example, for the Parikh vector above, we would have $\CountOf{a} = 1$. You have
already seen this in Section~\ref{sec:introduction:parikh}.

Parikh's theorem states that any context-free language has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars and~\cite{bounds} for bounds on its size). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \CountOf{a} =
\CountOf{b} = \CountOf{c} \land \CountOf{a} \geq 0$).

\contents{\item this means we can represent this as Presburger
\item only interested in regular languages}

\subsection{Generalised Parikh maps}\label{sec:generalised}

To generalise the logic of the Parikh map, we introduce a morphism $\Map:\: \Sigma^*
\to \Monoid$ where $\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative monoid where the following holds:
\begin{itemize}
  \item $\Map(\epsilon) = 0_{\Monoid}$
  \item $\Map(a \Concat b) = \Map(a) +_{\Monoid} \Map(b)$, where $\Concat$ is
  the string concatenation operation
\end{itemize}
